Regularized target encoding outperforms traditional methods in supervised machine learning with high cardinality features

### 2022-02-04 publication/manuscript.Rmd
make additions for second revision

While increasing sample size is usually considered the most important step to improve the predictive performance of a machine learning (ML) model, using effective feature engineering comes as a close second. One remaining challenge is how to handle high cardinality features -- categorical predictor variables with a high number of different levels but without any natural ordering.
While categorical variables with only a small number of possible levels can often be efficiently dealt with using standard techniques such as one-hot encoding, this approach becomes inefficient as the number of levels increases.
Despite this inefficiency, simpler strategies are often favored in practice because other methods are either not known, implementations are missing or because of a lack of trust due to missing validation studies.
Although domain knowledge can sometimes be used to reduce the number of theoretically relevant levels, finding strategies that work well on a large variety of problems is highly important for many applications as well as in automated ML [@thomas_2018; @thornton_2013; @feurer_2015]. Optimally, strategies should be model-agnostic because benchmarking encoding methods together with ML algorithms from different classes is often necessary for applications.
While a variety of strategies exist, there are very few benchmarks that can be used to decide which technique is expected to yield good predictive performance.
Furthermore, there has recently been increasing attention on scientific benchmark studies that compare different methods to provide a clearer picture in light of a large number of methods available to practitioners [@delgado; @bommert], as they can provide at least partial answers to such questions.
The goal of this study is to provide an overview of existing approaches for encoding categorical predictor variables and to study their effect on a model's predictive performance.
Following calls for neutral benchmark studies [@boulesteix_2017], which do not introduce a new method, thus reducing the risk of cherry picking methods [@dehghani2021benchmark] and reporting over-optimistic performance [@niesl2021overoptimism], we present a carefully designed experimental setting to discern the effect of encoding strategies and their interaction with different ML algorithms.

@@

Several small-scale studies have been previously conducted.
Those works did not yield conclusive results due to narrower scopes or not considering high cardinality variables.
One benchmark ($6$ datasets) on encoding high cardinality features (cardinality between $103$ and $9095$) in combination with gradient boosting has been published on the Kaggle forums [@prokopev_2018].
Different versions of target encoding are compared with indicator, integer, and frequency encoding.
They recommend combining the smoothed version of target encoding with 4- or 5-fold CV, never using simple target encoding and using indicator encoding only for small datasets.
Interestingly, frequency encoding did perform well in many cases.
@coors_2018 performed a benchmark ($12$ datasets) on encoding high cardinality features (maximum number of levels per dataset between 10 and 25847) when developing the automatic gradient boosting (autoxgboost) library [@thomas_2018]. They compared different variants of target encoding with integer and indicator encoding. In their benchmarks, target encoding only improved over target-agnostic methods on $2$ datasets, while it led to worse results on 4 datasets. For a smaller number of levels, indicator and integer encoding yielded similar results.
As those studies only consider gradient boosting and a limited amount of datasets, it is unclear whether results generalize to other datasets and ML algorithms.
A recent study ($15$ datasets) found good performance of target encoding, but they only investigated categorical features in regression settings [@seca_2021]. 
Several other publications studied encoding text data based on similarity [@cerda_2018; @cerda_2020] and employed indicator or target encoding as baselines.
Another line of work studies variable encodings employed within specific ML models.
@wright_2019 study treatments for categorical variables in random forests together with dummy and integer encoding (18 datasets, cardinality between $3$ and $38$), concluding that indicator and integer encoding perform subpar in comparison to methods that re-order levels according to the target variable.
@prokhorenkova_2018 compared their new CatBoost variant of target encoding to smoothed target encoding without CV, hold-out, and leave-one-out CV on $8$ datasets.
While the CatBoost method performed best, hold-out came second and target encoding without CV performed worst.
An overview of encoding techniques tailored towards neural networks (e.g. the widely adopted *entity embeddings* by @guo_2016) is provided in @hancock.
They present a survey of indicator- and embedding-based methods but no benchmark study.
In contrast, our work is the first to focus on high cardinality variables and techniques that are agnostic to the subsequent ML method.
We study this problem on a larger variety of datasets and settings.
\newline

@@

To further summarize, present meta rankings for each ML algorithm in Figure \ref{fig:rank-res}.
First, we defined an encoder relation within each dataset, based on corrected resample t-tests [@nadeau_2003].
An encoding was defined to beat another encoding if the one-sided p-value of the t-test was $< .05$.
This allowed us to compute a weak-order consensus ranking $R$ defined by the optimization problem:

@@

In an exploratory analysis, we tried to find clusters of datasets based on systematic patterns of encoder performance (independent of the employed ML algorithm).
We computed a partial-order consensus relation across ML algorithms for each dataset and hierarchically clustered the dataset consensus relations using the symmetric difference distance in combination with the complete linkage agglomeration method.
The resulting dendrogram is displayed in Figure \@ref(fig:dendrogram).
Although the cluster structure is somewhat ambiguous, roughly three clusters can be described:
The first 9 datasets from the top of the dendrogram are characterized by a medium to high number of levels, low performance of the remove condition (indicating the importance of high cardinality features) and clear performance advantage of target encoders.
For the next 13 datasets which contain the smallest number of levels, traditional encodings can compete with target encoding.
This biggest cluster also includes 9 datasets with zero distances, in which no significant performance differences could be observed between any encoding conditions (nor with the remove condition, indicating that high cardinality features are less informative).
The last two datasets with the highest number of levels formed a separate cluster, in which target encoding without strong regularization (impact, glmm-noCV) showed severe overfitting.
Note that clusters were not determined by problem type, again suggesting that encoder rankings are somewhat similar for regression and classification settings.\footnote{We tried to corroborate this observation by comparing meta-rankings between problem types. Unfortunately, problem type specific consensus relations did not converge (probably due to the reduced number of datasets) and were uninterpretable.}

@@

Several design decisions were necessary to make answering our research questions computationally feasible:
We used minimal tuning for our ML algorithms, as we were not interested in comparing their performance against each other.
This probably led to suboptimal performance for the GB, KNN, and SVM learners.
When interpreting our results, we assume that the encoder rankings are comparable when more extensive tuning is used.
This is plausible because the meta rankings between algorithms were highly stable.
We only used 5-fold-CV without repetitions to estimate predictive performance, but due to a small variance between folds, we could confidently detect performance differences of encoders for many datasets.
Because we are interested in model-agnostic methods that can be combined with any supervised ML algorithm, we did not investigate recent model-specific strategies [@guo_2016; @prokhorenkova_2018].
We do not differentiate between nominal and ordinal categorical features:
Most publicly available datasets simply do not contain any high cardinal ordinal features.
For many applications, ordinal feature information is not available as metadata which makes it less relevant for some applications like automatic ML.
Thus, our benchmark does not include specific ordinal encoding methods like Helmert and Polynomial contrasts [@chambers_1992] or ordinal penalization methods [@tutz_2016].
We only investigate traditional categorical variables and do not extend our analysis to multi-categorical variables or text-strings [@cerda_2018; @cerda_2020], where other encoding techniques can harvest additional information.
We only investigate univariate encodings, i.e. we always encode each variable separately.
Levels with comparable main effects can not be distinguished based on the transformed feature, which prevents the consecutive ML algorithm from learning interactions for specific levels.
Methods that jointly encode features and therefore leverage correlational structures between features are an interesting avenue for future research:
Recent approaches use target-based encoding strategies to learn a vector-valued representation of a level, similar to neural network embeddings [@guo_2016;@rodriguez_2018].
Those should be compared to classical approaches from the optimal scaling literature [e.g. @deleeuw_1976;@young_1976] or the subsequent *aspect* framework [@mair_2010], which are unfortunately unfamiliar to many ML researchers.

@@

This benchmark study compared the predictive performance of a variety of strategies to encode categorical features with a high number of unordered levels for different supervised ML algorithms.
Regularized versions of target encoding, which uses predictions of the target variable as numeric feature values, performed better than traditional strategies like integer or indicator encoding.
Most effective, with a consistently superior performance across Ml algorithms and datasets, was a target encoder which combines simple generalized linear mixed models with cross-validation and does not require hyperparameter tuning.

>>>>

While increasing sample size is usually considered the most important step to improve the predictive performance of a machine learning (ML) model, using effective feature engineering comes as a close second. One remaining challenge is how to handle high cardinality features -- categorical predictor variables with a high number of different levels but without any natural ordering.
While categorical variables with only a small number of possible levels can often be efficiently dealt with using standard techniques such as one-hot encoding, this approach becomes inefficient as the number of levels increases.
Despite this inefficiency, simpler strategies are often favored in practice because other methods are either not known, implementations are missing or because of a lack of trust due to missing validation studies.
Although domain knowledge can sometimes be used to reduce the number of theoretically relevant levels, finding strategies that work well on a large variety of problems is highly important for many applications as well as in automated ML [@thomas_2018; @thornton_2013; @feurer_2015]. Optimally, strategies should be model-agnostic because benchmarking encoding methods together with ML algorithms from different classes is often necessary for applications.
While a variety of strategies exist, there are very few benchmarks that can be used to decide which technique is expected to yield good predictive performance.
Furthermore, there has recently been increasing attention on scientific benchmark studies that compare different methods to provide a clearer picture in light of a large number of methods available to practitioners [@delgado; @bommert], as they can provide at least partial answers to such questions.
The goal of this study is to provide an overview of existing approaches for encoding categorical predictor variables and to study their effect on a model's predictive performance.
Following calls in the computational statistics community for neutral benchmark studies [@boulesteix_2017], which do not introduce a new method, thus reducing the risk of cherry picking methods [@dehghani2021benchmark] and reporting over-optimistic performance [@niesl2021overoptimism], we present a carefully designed experimental setting to discern the effect of encoding strategies and their interaction with different ML algorithms.

@@

Several small-scale studies have been previously conducted.
Those works did not yield conclusive results due to narrower scopes or not considering high cardinality variables.
One benchmark ($6$ datasets) on encoding high cardinality features (cardinality between $103$ and $9095$) in combination with gradient boosting has been published on the Kaggle forums [@prokopev_2018].
Different versions of target encoding are compared with indicator, integer, and frequency encoding.
They recommend combining the smoothed version of target encoding with 4- or 5-fold CV, never using simple target encoding and using indicator encoding only for small datasets.
Interestingly, frequency encoding did perform well in many cases.
@coors_2018 performed a benchmark ($12$ datasets) on encoding high cardinality features (maximum number of levels per dataset between 10 and 25847) when developing the automatic gradient boosting (autoxgboost) library [@thomas_2018]. They compared different variants of target encoding with integer and indicator encoding. In their benchmarks, target encoding only improved over target-agnostic methods on $2$ datasets, while it led to worse results on 4 datasets. For a smaller number of levels, indicator and integer encoding yielded similar results.
As those studies only consider gradient boosting and a limited amount of datasets, it is unclear whether results generalize to other datasets and ML algorithms.
A recent study ($15$ datasets) found good performance of target encoding, but they only investigated categorical features in regression settings [@seca_2021]. 
Several other publications studied encoding text data based on similarity [@cerda_2018; @cerda_2020] and employed indicator or target encoding as baselines.
Another line of work studies variable encodings employed within specific ML models.
@wright_2019 study treatments for categorical variables in random forests together with dummy and integer encoding (18 datasets, cardinality between $3$ and $38$), concluding that indicator and integer encoding perform subpar in comparison to methods that re-order levels according to the target variable.
@prokhorenkova_2018 compared their new CatBoost variant of target encoding to smoothed target encoding without CV, hold-out, and leave-one-out CV on $8$ datasets.
While the CatBoost method performed best, hold-out came second and target encoding without CV performed worst.
An overview of encoding techniques tailored towards neural networks [e.g. the widely adopted *entity embeddings* by @guo_2016] is provided in @hancock.
They present a survey of indicator- and embedding-based methods but no benchmark study.
In contrast, our work is the first to focus on high cardinality variables and techniques that are agnostic to the subsequent ML method.
We study this problem on a larger variety of datasets and settings.
\newline

@@

To further analyze our results, we used statistical inference methods inspired by the benchmark community in computational statistics [e.g., @hothorn_2005]. 
First, we present meta rankings for each ML algorithm in Figure \ref{fig:rank-res}:
We defined an encoder relation within each dataset, based on corrected resample t-tests [@nadeau_2003].
An encoding was defined to beat another encoding if the one-sided p-value of the t-test was $< .05$.
This allowed us to compute a weak-order consensus ranking $R$ defined by the optimization problem:

@@

In a second exploratory analysis, we tried to find clusters of datasets based on systematic patterns of encoder performance (independent of the employed ML algorithm).
We computed a partial-order consensus relation across ML algorithms for each dataset and hierarchically clustered the dataset consensus relations using the symmetric difference distance in combination with the complete linkage agglomeration method.
The resulting dendrogram is displayed in Figure \@ref(fig:dendrogram).
Although the cluster structure is somewhat ambiguous, roughly three clusters can be described:
The first 9 datasets from the top of the dendrogram are characterized by a medium to high number of levels, low performance of the remove condition (indicating the importance of high cardinality features) and clear performance advantage of target encoders.
For the next 13 datasets which contain the smallest number of levels, traditional encodings can compete with target encoding.
This biggest cluster also includes 9 datasets with zero distances, in which no significant performance differences could be observed between any encoding conditions (nor with the remove condition, indicating that high cardinality features are less informative).
The last two datasets with the highest number of levels formed a separate cluster, in which target encoding without strong regularization (impact, glmm-noCV) showed severe overfitting.
Note that clusters were not determined by problem type, again suggesting that encoder rankings are somewhat similar for regression and classification settings.\footnote{We tried to corroborate this observation by comparing meta-rankings between problem types. Unfortunately, problem type specific consensus relations did not converge (probably due to the reduced number of datasets) and were uninterpretable.}

@@

Several design decisions were necessary to make answering our research questions computationally feasible:
We used minimal tuning for our ML algorithms, as we were not interested in comparing their performance against each other.
This probably led to suboptimal performance for the GB, KNN, and SVM learners.
When interpreting our results, we assume that the encoder rankings are comparable when more extensive tuning is used.
This is plausible because the meta rankings between algorithms were highly stable.
We only used 5-fold-CV without repetitions to estimate predictive performance, but due to a small variance between folds, we could confidently detect performance differences of encoders for many datasets.
Because we are interested in model-agnostic methods that can be combined with any supervised ML algorithm, we did not investigate recent model-specific strategies [@guo_2016; @prokhorenkova_2018].
We did not include deep neural networks in our benchmark because they are still rarely applied in the statistical learning community and more dominant in computer science. 
We do not differentiate between nominal and ordinal categorical features:
Most publicly available datasets simply do not contain any high cardinal ordinal features.
For many applications, ordinal feature information is not available as metadata which makes it less relevant for some applications like automatic ML.
Thus, our benchmark does not include specific ordinal encoding methods like Helmert and Polynomial contrasts [@chambers_1992] or ordinal penalization methods [@tutz_2016].
We only investigate traditional categorical variables and do not extend our analysis to multi-categorical variables or text-strings [@cerda_2018; @cerda_2020], where other encoding techniques can harvest additional information.
We only investigate univariate encodings, i.e. we always encode each variable separately.
Levels with comparable main effects can not be distinguished based on the transformed feature, which prevents the consecutive ML algorithm from learning interactions for specific levels.
Methods that jointly encode features and therefore leverage correlational structures between features are an interesting avenue for future research:
Recent approaches use target-based encoding strategies to learn a vector-valued representation of a level, similar to neural network embeddings [@guo_2016;@rodriguez_2018].
Those should be compared to classical approaches from the optimal scaling literature [e.g. @deleeuw_1976;@young_1976] or the subsequent *aspect* framework [@mair_2010], which are unfortunately unfamiliar to many ML researchers.

@@

This benchmark study compared the predictive performance of a variety of strategies to encode categorical features with a high number of unordered levels for different supervised ML algorithms.
Regularized versions of target encoding, which uses predictions of the target variable as numeric feature values, performed better than traditional strategies like integer or indicator encoding.
Most effective, with a consistently superior performance across Ml algorithms and datasets, was a target encoder which combines simple generalized linear mixed models with cross-validation and does not require hyperparameter tuning.
GLMMs are a major workhorse in applied statistics but not well understood and often neglected by the ML community.
Refining target encoders that use statistical models for more efficient regularization and studying their theoretical properties could be a valuable research topic for the computational statistics community.
